---
title: "Exploratory Analysis"
output: html_document
---

## Project title

>**Understanding Climate Change through Data Analysis**

### Authors

>Natnael Tekie, 2122558, Ntekie@uw.edu

### Date

>April 30<sup>th</sup>, Winter 2024

### Abstract

>This project aims to analyze various aspects of climate change through data exploration and analysis. 
By examining datasets related to climate indicators, greenhouse gas emissions, and environmental impacts, 
we seek to uncover patterns, trends, and correlations to deepen our understanding of climate change dynamics.

### Keywords

3-5 keywords that summarize your project

>Climate Change, Data Analysis, Environmental Science, Greenhouse Gas Emissions, Sustainability

### Introduction 
>This project seeks to delve into the multifaceted issue of climate change by posing several research questions:

>- What are the trends in global temperature, sea level rise, and other climate indicators over the past century?
>- How do human activities, such as industrialization and deforestation, contribute to greenhouse gas emissions?
>- What are the regional and global impacts of climate change on ecosystems, biodiversity, and human societies?
>- What mitigation and adaptation strategies are being implemented to address climate change, and what are their effectiveness and challenges?</br>
<p>

>These questions are motivated by the urgency of understanding and addressing climate change, 
a defining challenge of the 21<sup>st</sup> century with profound implications 
for the environment, economy, and society.

### Related Work  

>Climate change has been extensively studied across various disciplines, with numerous 
research papers, reports, and initiatives dedicated to understanding its causes, impacts, 
and solutions. For example, the Intergovernmental Panel on Climate Change (Hansen et al., 2018) 
provides comprehensive assessments of the scientific basis of climate change and its 
potential impacts. Additionally, academic studies such as those by Le Quéré et al. (2020) 
offer insights into the latest trends in global temperature rise. Furthermore, organizations 
like NASA and NOAA provide valuable datasets and analyses on climate indicators and 
environmental changes.

### The Dataset

> **Where did you find the data?** 
*Please include a link to the data source*<br/>
This dataset in use was obtained from Kaggle, the largest online data science repository. 
https://www.kaggle.com/datasets/berkeleyearth/climate-change-earth-surface-temperature-data

> **Who collected the data?**  
The data was collected by Berkelye Earth, California-based independent 501 non-profit focused on land temperature data analysis for climate science. 
Additionally, the Kaggle team played a major collaboration role in the data collection process. <br/>

> **How was the data collected or generated?**  
The Berkeley Earth Surface Temperature Study combines 1.6 billion temperature reports from 16 pre-existing archives. It is nicely packaged and allows for slicing into interesting subsets (for example by country). They publish the source data and the code for the transformations they applied.

> **Why was the data collected?**  
Berkeley collectes and publishes scientific and data analysis data to help deliver and communicate their findings to the broadest possible audience.

```{r summary_info, echo=FALSE}
# Load the data
temperature_data <- read.csv("Datasets/GlobalTemperatures.csv", stringsAsFactors = FALSE)
```

> **How many observations (rows) are in your data?**  
```{r summary_info echo=FALSE}
r nrow(temperature_data) 
```

> **How many features (columns) are in the data?**  
```{r summary_info echo=FALSE}
r ncol(temperature_data) 
```

> **What, if any, ethical questions or questions of power do you need to consider when working with this data?**  
When using temperature data from Berkeley Earth, ethical considerations include evaluating the accuracy of instruments used for data collection and ensuring transparency regarding calibration and potential biases. 

> **What are possible limitations or problems with this data?   (at least 200 words)** <br/>
Berkeley Earth data contains n/a values, posing a significant challenge due to their potential to distort analyses and undermine data integrity. Addressing these n/a values is crucial, as they can stem from equipment malfunctions or transmission errors, compromising the reliability of findings. <br/>
Conversely, missing data within the dataset presents another substantial challenge, as it hampers comprehensive trend assessments and undermines the completeness of analyses. This issue is exacerbated by gaps in monitoring infrastructure or discontinuities, necessitating robust strategies like interpolation or sensitivity analysis to mitigate its impact. <br/>
Therefore, while Berkeley Earth data offers valuable insights into climate trends, addressing n/a values and missing data is paramount for ensuring the accuracy and reliability of research findings and subsequent policy decisions.

### Implications

>- Policymakers can use the insights to inform evidence-based climate policies, set emissions reduction targets, and prioritize adaptation measures.
>- Scientists and researchers can leverage the findings to advance climate science, develop predictive models, and identify areas for further investigation.
>- Civil society organizations and advocacy groups can use the data to raise awareness, mobilize action, and hold governments and corporations accountable for their environmental commitments.

### Limitations & Challenges
>- Complexity of climate data analysis: Climate datasets are often vast, heterogeneous, and complex, requiring advanced statistical and computational techniques for analysis and interpretation.
>- Uncertainties and variability: Climate data may contain inherent uncertainties and natural variability, making it challenging to attribute specific events or trends solely to human-induced climate change.
>- Interdisciplinary nature: Climate change intersects with various disciplines, including environmental science, economics, sociology, and policy studies, requiring interdisciplinary collaboration and expertise.

### Summary Information
>The exploratory analysis of global land temperature datasets reveals interesting 
patterns across various geographical regions. The dataset of average temperatures 
by city shows a mean temperature of approximately 16.727°C. In the country dataset,
the median average temperature is around 20.901°C. The highest average temperature 
recorded in a major city is 38.283°C, observed in Baghdad, Iraq. In contrast, the 
lowest temperature recorded in a state is -45.389°C, found in Sakha, Russia. 
Additionally, the standard deviation of the global land and ocean average 
temperature is 1.274°C, indicating the variability of temperature measurements 
across different locations. These statistics provide a comprehensive overview of 
temperature variations and highlight regions with extreme temperature records.


### Table
#### Aggregated Data for Cities

| Country | City        | avg_temp | median_temp | min_temp | max_temp | avg_uncertainty |
|---------|-------------|----------|-------------|----------|----------|-----------------|
| Djibouti| Jibuti      | 29.2     | 28.8        | 23.2     | 35.8     | 0.923           |
| Sudan   | Khartoum    | 29.1     | 30.3        | 18.5     | 35.7     | 0.891           |
| Sudan   | Umm Durman  | 29.1     | 30.3        | 18.5     | 35.7     | 0.891           |
| Niger   | Niamey      | 29.1     | 29.0        | 21.3     | 35.6     | 0.779           |
| Sudan   | Kassala     | 28.9     | 29.1        | 21.5     | 35.5     | 0.861           |
| Mali    | Mopti       | 28.8     | 29.6        | 20.1     | 35.6     | 0.890           |
| Sudan   | Kusti       | 28.6     | 29.0        | 19.6     | 34.8     | 0.807           |
| Sudan   | Rabak       | 28.6     | 29.0        | 19.6     | 34.8     | 0.807           |
| India   | Alandur     | 28.4     | 28.7        | 21.6     | 34.6     | 0.999           |
| India   | Ambattur    | 28.4     | 28.7        | 21.6     | 34.6     | 0.999           |

#### Aggregated Data for Countries

| Country             | avg_temp | median_temp | min_temp | max_temp | avg_uncertainty |
|---------------------|----------|-------------|----------|----------|-----------------|
| Djibouti            | 28.8     | 28.4        | 23.0     | 35.2     | 0.885           |
| Mali                | 28.4     | 30.0        | 19.1     | 35.3     | 0.775           |
| Burkina Faso        | 28.1     | 27.7        | 22.9     | 33.7     | 0.698           |
| Senegal             | 28.0     | 28.2        | 21.5     | 32.9     | 0.710           |
| Aruba               | 27.9     | 28.1        | 24.1     | 30.8     | 0.645           |
| United Arab Emirates| 27.7     | 28.3        | 15.7     | 37.8     | 0.811           |
| Mauritania          | 27.6     | 28.9        | 17.3     | 34.7     | 0.851           |
| Gambia              | 27.5     | 27.7        | 21.6     | 31.7     | 0.725           |
| Niger               | 27.5     | 29.3        | 16.2     | 35.0     | 0.865           |
| Curaçao             | 27.4     | 27.6        | 23.7     | 30.0     | 0.645           |

#### Aggregated Data for States

| Country | State               | avg_temp | median_temp | min_temp | max_temp | avg_uncertainty |
|---------|---------------------|----------|-------------|----------|----------|-----------------|
| India   | Puducherry          | 27.3     | 27.4        | 22.2     | 31.9     | 0.849           |
| India   | Andaman And Nicobar | 27.2     | 27.3        | 23.3     | 30.2     | 0.855           |
| India   | Andhra Pradesh      | 27.1     | 27.3        | 19.9     | 34.4     | 0.891           |
| Brazil  | Rio Grande Do Norte | 27.0     | 27.1        | 24.2     | 29.5     | 0.686           |
| India   | Tamil Nadu          | 26.7     | 26.9        | 21.7     | 31.2     | 0.851           |
| India   | Kerala              | 26.3     | 26.0        | 23.1     | 30.3     | 0.822           |
| India   | Gujarat             | 26.3     | 27.2        | 16.7     | 34.0     | 0.868           |
| India   | Daman And Diu       | 26.3     | 27.0        | 19.2     | 31.4     | 0.894           |
| Brazil  | Amazonas            | 26.1     | 26.1        | 24.0     | 28.2     | 0.594           |
| India   | Maharashtra         | 25.9     | 25.9        | 18.1     | 33.8     | 0.866           |

#### Description
1. **City Data**:
   - **Purpose**: This table aggregates temperature data by city, revealing the average, median, minimum,
     and maximum temperatures, along with the average uncertainty.
   - **Insight**: It highlights which cities have the highest and lowest average temperatures. For example,
     cities in Djibouti and Sudan have some of the highest average temperatures, whereas Indian cities
     like Alandur and Ambattur also have high averages but with higher uncertainty.

2. **Country Data**:
   - **Purpose**: This table aggregates temperature data by country, showing the average, median, minimum,
     and maximum temperatures, along with average uncertainty.
   - **Insight**: Countries like Djibouti and Mali consistently have high average temperatures. The data
     also reveals temperature variation within countries, useful for understanding regional climate
     differences.

3. **State Data**:
   - **Purpose**: This table aggregates temperature data by state, providing insights into temperature
     variations within countries.
   - **Insight**: States in India and Brazil are prominently featured, showing significant temperature
     ranges. This data is crucial for local climate studies and for understanding how temperature changes
     within a country.

### Why Include These Tables?

- **Comprehensive Overview**: These tables offer a clear and concise summary of temperature data across
  different geographical levels, providing a broad perspective on global temperature variations.
- **Comparative Analysis**: By presenting data for cities, countries, and states, it allows for comparison
  between different regions, helping to identify patterns and anomalies.
- **Data-Driven Decisions**: This aggregated information is valuable for researchers, policymakers, and
  analysts to make informed decisions based on temperature trends and uncertainties.
- **Highlight Extremes**: Including cities and states with the highest and lowest temperatures helps to
  understand extreme climate conditions and their potential impacts.

### Chart 1
```{r, echo = FALSE, code = readLines("FirstChart.R")}
library(ggplot2)

# Chart 1: Average Temperature by Country
country_chart <- ggplot(country_table, aes(x = reorder(Country, avg_temp), y = avg_temp)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Average Temperature by Country", x = "Country", y = "Average Temperature") +
  theme_minimal()

print(country_chart)

```

**Why Chosen**: 
>This chart effectively displays the average temperature by country, allowing for a clear
visual comparison of temperature variations across different countries.

**Information Revealed**: 
>It highlights countries with the highest and lowest average temperatures, providing
insight into global temperature distribution. This is particularly useful for identifying regions with
extreme climates and understanding overall global temperature trends.



### Chart 2
```{r, echo = FALSE, code = readLines("SecondChart.R")}
library(ggplot2)

# Chart 2: Median Temperature by State
state_chart <- ggplot(state_table, aes(x = reorder(State, median_temp), y = median_temp)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  labs(title = "Median Temperature by State", x = "State", y = "Median Temperature") +
  theme_minimal()

print(state_chart)

```

**Why Chosen:** 
>This chart presents the median temperature by state, offering a focused view on temperature
variations within countries.

**Information Revealed:** 
>It shows which states experience the highest and lowest median temperatures,
highlighting regional climate differences within countries. This can help in understanding local climate
patterns and the range of temperatures within larger geographic areas.

### Chart 3
```{r, echo = FALSE, code = readLines("ThirdChart.R")}
library(ggplot2)
temperature_by_country$dt <- as.Date(temperature_by_country$dt)

# Plot the temperature distribution over time
temperature_over_time <- ggplot(temperature_by_country, aes(x = dt, y = AverageTemperature)) +
  geom_line() +
  labs(title = "Temperature Distribution Over Time",
       x = "Date",
       y = "Average Temperature") +
  theme_minimal()

print(temperature_over_time)
```

**Why Chosen:**
>This chart visualizes the distribution of average temperatures over time. 
plotting a line chart showing how average temperatures have changed over the years.

**Information Revealed:**
>By observing the trend of temperature distribution over time, we can identify 
long-term climate patterns such as global warming or cooling trends.
Significant deviations from the expected temperature range can be easily spotted

### References 
>-  Hansen, J., Sato, M., & Ruedy, R. (2019). Global temperature update through 2018. GISS.
>-  Le Quéré, C., Jackson, R. B., Jones, M. W., Smith, A. J., Abernethy, S., Andrew, R. M., ... & Canadell, J. G. (2020). Temporary reduction in daily global CO2 emissions during the COVID-19 forced confinement. Nature Climate Change, 10(7), 647-653.
